---
title: "Final EDA Assignment"
author: "Pooja Chopra"
date: "16 June 2019"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

In this EDA project, I have used the Gapminder Dataset for analysis. First We will be following the epicycle analysis in R which we have learned during this course. We will start with setting up questions.

##  Setting  QUESTIONS

Q1 - How per capita income in SouthAsia region has trended over the Years? Which country per capita income increases the most in south asia region?
Q2 - What was the percent growth in world GDP per capita in 2015 as compared to 1800?
Q3 - Does life expectancy increase with increase in income? Is there enough evidence that high income does not lead to high life expectancy for a country/region?


Let us start with analysis, results and their interpretations:

## Importing Required Libraries

```{r Libraries, echo=FALSE, message=FALSE, warning=FALSE}

library(readr)
library(tidyverse)
library(mice)
library(lsr)
library(naniar)
library(ggplot2)
library(DataExplorer)
library(outliers)
library(dlookr)
library(matrixStats)
library(dplyr)
library(Hmisc)
library(DMwR)
library(mice)
library(cluster)
library(factoextra)
```

##Importing raw data
```{r Importing data, echo=FALSE, message=FALSE, warning=FALSE}

setwd("C:/Users/pooja/Desktop/HU Semester/HU Semester 2/506/Group project/506-Final-Project/Final Assignment")

Gapminder_Raw <- read_csv("gapminder.csv",col_types = 'fnnfif')

```

## Data Description

```{r Data Description, echo=FALSE, message=FALSE, warning=FALSE , results='hide'}

nrow(Gapminder_Raw)       # Number of rows: 41284
ncol(Gapminder_Raw)      # Number of columns: 6
names(Gapminder_Raw)   # Display variable name

```
Region/Country describes the continent and country the Life expectancy is measured
Population, Income, and Life expectance is provided for each country from years 1800 to 2015

Gapminder dataset have 41284 observation with 6 variables. 
Type of Variable Country is categorical variable with 197 Levels,Region is categorical variable with 6 Levels,Year,Life and Income are continous variable and Population are descrete variable

##Dispersion of Data

```{r Dispersion of Data, echo=FALSE, message=FALSE, warning=FALSE,results='hide' }

Range_Year <- max(Gapminder_Raw$Year)-min(Gapminder_Raw$Year)

Range_life  <- max(Gapminder_Raw$life)-min(Gapminder_Raw$life)

Mean_adj_deviation_year <- aad(Gapminder_Raw$Year)
Mean_adj_deviation_life <- aad(Gapminder_Raw$life)

#Number of Unique Countries
length(unique(Gapminder_Raw$Country))

```

In our dataset, data has been gathered for over 215 years starting from 1800 up till 2015. It includes the population of 197 countries spread across 6 regions changed over time. The life expectacy ranges from 1 to 83%.

## Data Wrangling  


```{r Handling missing data, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}


#Finding NA's in Dataset

md.pattern(Gapminder_Raw)

plot_missing(Gapminder_Raw)

#Filling NA value in Population column


Gapminder_Raw$population <- as.numeric(gsub(",","",Gapminder_Raw$population))
Gapminder_Filtered <- Gapminder_Raw

Gapminder_Filtered <- Gapminder_Raw %>% fill(population)

min(Gapminder_Filtered$population)


# Handling NA values in Income columns 

Gapminder_NAincome <- Gapminder_Filtered[is.na(Gapminder_Filtered$income),]
summary(Gapminder_NAincome)



```
In our data set we have 25817(62.54 %) missing values in population and 2341(5.67%) missing values in Income. 
In our dataset population is calculated every ten years, we have filled missing values same as calcualted at the start of the decade.
In our dataset Missing values of income are for multipe countries and it is observed that Values were never calculated from the start so it is not possible to handle missing value of Income hence we are eliminating observations where we have NA income. 

##Preprocessing
```{r Outliers , echo=FALSE, message=FALSE, warning=FALSE }

outlier_Life = outlier(Gapminder_Filtered$life,logical=TRUE)
outlier_population = outlier(Gapminder_Filtered$population,logical=TRUE)

find_outlier_Population = which(outlier_population==TRUE,arr.ind=TRUE)

find_outlier_life = which(outlier_Life==TRUE,arr.ind=TRUE)

## Log for population in dataset

Gm_final <- Gapminder_Filtered

Gm_final$population <- log(Gapminder_Filtered$population)

```
We used the function outliers only on the numeric columns which are Life ,Population . we find few outliers in out data which can be ignored as it will not impact our analysis. We have taken log for population column for further analysis.

##Data Exploration
```{r life_expectancy , echo=FALSE, message=FALSE, warning=FALSE }

hist(Gm_final$life)

Life_plot <- ggplot(Gm_final, aes(x = life, y = income)) + geom_point()+geom_smooth(method = "lm") + scale_x_continuous(labels = scales::comma) + labs(title="Imapct of income on Life expectancy", x="life", y="income") 

print(Life_plot)

```

Looking at the histogram , Life expectancy trends to increase after 20 till 35. In other plot we can conclude that life expectancy is increasing regardless of growth in income but EXpectacy level grows above 60 when the income levels increase. 


## Regionwise ANalysis

```{r Region,Life , echo=FALSE, message=FALSE, warning=FALSE }

#Regionwise analysis

median_Life_region<- aggregate(life ~ region, Gm_final, median)

mean_Life_region<- aggregate(life ~ region, Gm_final, mean)

mean_lifeexp<-round(mean(Gm_final$life), digits=3) #whole world mean 
medn_lifeexp<-round(median(Gm_final$life), digits=3)

Gm_final %>% group_by(region) %>% summarise(mean=mean(life) , median=median(life))


## Life expectancy trending over region

ggplot(Gm_final, aes(x = region, y = life))  + geom_boxplot(outlier.colour = "red") +  
geom_jitter(position = position_jitter(width = 0.1, height = 0), alpha = 1/4) + labs(title="Life expectancy trending over region", x="Region", y="Life") 



```
Looking at the Life expectacy trending over region Graph, Europe region has high level of Life Expectacy as compared to our Regions , America and East Asia pacific is also high levl of life expectact. 

## income Study

```{r income study,echo=FALSE, message=FALSE, warning=FALSE }


## Income with Region

Gm_final %>% group_by(region) %>% summarise(mean = mean(income,na.rm = TRUE),
median = median(income,na.rm = TRUE), min = min(income,na.rm = TRUE), max = max(income,na.rm = TRUE))

gmCONT <- Gm_final %>% group_by(region)

ggplot(data = gmCONT, mapping = aes(x = region, 
  y = income)) + geom_boxplot() + 
  ggtitle("Distribution compare for Income")



## Income percentage increase from 1800 to 2015 

World_GDP<- filter(Gm_final, Year %in% c('2015','1800'))
World_GDP_2015 <- group_by(World_GDP, Year)
World_GDP_summarise<- summarise(World_GDP_2015, avg_growth_Percap = mean(income, na.rm = TRUE))


## How per capita income in SouthAsia region has trended over the Years? Which country per capita income increases the most in south asia region?

SA_per_capita <- subset(Gm_final, Gm_final$region=='South Asia')

SA <- ggplot(data=SA_per_capita) +
  geom_point(aes(x=Year, y= income,color= Country)) +
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=.025))+ggtitle("Income Trend over years fro South Asia Region")
SA


```

"The Percent growth (or decline) in GDP per capita in 2015 for the world was 18.62%".
From distribution compare for income Plot we observed that , Europe income is substantially higher than all regions.  If we look into Income and Region data for south asia region we observed that Per capita income is tend to increase over the Year and Among all countries Maldives per capita increases the most in south Asia region till 2014.


## Population

```{r Population,Region , echo=FALSE, message=FALSE, warning=FALSE }


Gm_final %>% group_by(region) %>% summarise(max=max(population) , min=min(population))


Gm_final %>% group_by(region)%>% summarise(Total_population = sum(population))

Gm_final %>% group_by(region)%>% summarise(Total_population = sum(population), Income=sum(income,na.rm=TRUE))

ggplot(Gm_final, aes(x = region, y = population))  + geom_boxplot(outlier.colour = "red") +  
geom_jitter(position = position_jitter(width = 0.1, height = 0), alpha = 1/4) + labs(title="Regionwise Population Distribution of ", x="Region", y="Population") 

```

With this analysis we observed that population of Asia pacific region is higher than other regions. If we compare Regionwise population with income Europe & Central Asia has highest population with highest income  we can say that increase in population increase income.

## Country

```{r Comparing Population for China and America , echo=FALSE, message=FALSE, warning=FALSE }

population_china <- Gm_final %>% filter(Country=='China')

population_america <- Gm_final %>% filter(Country=='Virgin Islands (U.S.)')

ggplot(data=population_china) +
  geom_point(aes(x=population, y= Year,color= Country)) +
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=.025))+ggtitle("Yearly China Population")

ggplot(data=population_america) +
  geom_point(aes(x=population, y= Year,color= Country)) +
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=.025))+ggtitle("Yearly America Population")

```
Here we can observe that population for both countries China and Virginia island increase after 1950's

## Summary

In this Analysis we observed that The mean of Life Expectancy  for *Sub-Saharan Africa* region is lower than all other regions, while Europe & Central Asia is highest. *Q1*If we look into Income and Region data for south asia region we observed that Per capita income is tend to increase over the Years and Maldives per capita increases the most in south Asia region till 2015. *Q2* paste0("The Percent growth (or decline) in GDP per capita in 2015 for the world was ", round(World_GDP_summarise[2,2]/World_GDP_summarise[1,2], 2),"%")
Life expectancy in the Gapminder dataset is higher in the range of 25-35 years. Though it increase regardless of income *Q3* the income levels rises higher,life expectancy grows above 60 .With this analysis of population with region we observed that population of Asia pacific region is higher than other regions.  If we compare Regionwise population with income Europe & Central Asia has highest population with highest income  we can say that increase in population increase income.

## Clustering 


```{r kmeans, echo=FALSE, message=FALSE, warning=FALSE ,results='hide'}


# Kmeans

# filtering year 2015
gm_prepd_omited <- Gm_final[Gm_final$Year==2015,]

# Omiting NA and Scaling
gm_prepd_omited <- na.omit(gm_prepd_omited)
gm_prepd_scaled <- gm_prepd_omited
gm_prepd_scaled[3:5] <- scale(gm_prepd_omited[3:5])

# creating a subset with numeric values only
gm_clus_sub <- gm_prepd_omited[3:5] 

# calculating the Optimal number of clusters
set.seed(123)
fviz_nbclust(gm_clus_sub, kmeans, method = "silhouette")

# creating the cluster
k_cluster <- kmeans(gm_clus_sub, centers = 2, nstart = 25)

# adding cluster back to the dataset
gm_kmean_final <- gm_prepd_omited %>% mutate(Cluster = k_cluster$cluster)

gm_kmean_final$Country <- as.character(gm_kmean_final$Country)
gm_kmean_final$Cluster <- factor(gm_kmean_final$Cluster)

gm_kmean_final %>%
  as_tibble() %>%
  ggplot(aes(income, life, color = Cluster, label = Country)) +
  geom_text()+
  ggtitle("Life v/s Income - Kmeans Cluster")

```

In Kmeans Cluster the countries are more or less divided based on the life expectacy in 2015. The cutoff seems to be near 70 years of life expectancy. This also points to an interesting fact that richer countries don't actually live longer than poorer countries.


```{r hierar, echo=FALSE, message=FALSE, warning=FALSE ,results='hide'}

# Dissimilarity matrix
new <- dist(gm_clus_sub, method = "euclidean")

# Hierarchical clustering using Complete Linkage
hc1 <- hclust(new, method = "complete" )

# Cut tree into 4 groups
sub_grp <- cutree(hc1, k = 2)

# adding back to data
gm_hc_final <- gm_prepd_omited %>%  mutate(Cluster = sub_grp)
gm_hc_final$Cluster <- factor(gm_hc_final$Cluster)

gm_hc_final %>%
  as_tibble() %>%
  ggplot(aes(income, life, color = Cluster, label = Country)) +
  geom_text()+
  ggtitle("Life v/s Income - Hierarchical Cluster")


```


We observe results with the cluster being separated long life variable but here the cutoff seems to be slightly lower than 70 years.




